\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology

### 2017-07-10 mers-structure.tex
Incorporate @maxbiostat changes/comments.

@@

% symbols
\newcommand{\chiSq}{\chi^{2}_{df}} %LM: ancient DNA? =P
\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
\newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
\newcommand{\dspr}{d_\mathrm{SPR}}

@@

\date{\today}

@@

\section*{Introduction}
%% Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels. % AR - a virus isn't an infection and not all infections in the Arabian Peninsula
Middle East Respiratory Syndrome Coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans. % ? something like this ?
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
Whilst contact with camels is often reported, patients may not recall contact with \lmc{other?} livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
Epidemiological modeling approaches have been used to look at traditional data sources of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches \lmc{why, specifically?}.
The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Often only sequence data can \lmc{help} pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from hospital outbreaks \lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \lmc{how about a citation here?}.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.

%Unlike discrete trait analyses (DTA, or ``mugration'' models) %% AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
%that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) [cite here, Philippe's bat rabies and Sam's avian flu?, others?], population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using Monte Carlo simulations we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.

@@

The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we identify \lmc{estimate?} a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows \lmc{is ``allow'' the best word here? I understand you' rather not use ``biases'' or ``misleads'', though...} the model to consider humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly \lmc{maybe ``non-uniformly'' is redundant here} sampled human cases and are robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike the most basic \lmc{simpler?} ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, structured population models possess the ability to estimate migration times.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{MERS-CoV seasonality.}
A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Grey area spanning October to February indicates the camel calving season.
B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (Figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
	}
	\label{seasonality}
\end{figure}

@@

Structured coalescent approaches clearly show humans to be a terminal host %\tbc{again, I don't know if this is the right term or not}
for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides expectations\lmc{insights?} of epidemic behavior.
The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by a dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}).
We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
First, we simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from index cases will be sampled exceedingly rarely.
We record the $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses (left-hand panels in Figure \ref{ABC}).

We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%\lmc{again?} of simulations fit for bias levels of 2.0 and 3.0, respectively.
This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate a higher marginal likelihood for bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).\lmc{I don't follow this argument...}
Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).

Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{ABC}).
The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
Our results suggest a large number of unobserved MERS index cases.\lmc{does the definition of ``index case'' allow for ``multiple''?}
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In those sets of analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
	\caption{\textbf{Results of Monte Carlo simulations.}
Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias \lmc{make sequences from large clusters more likely to be picked} ).
Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
Darker shade of grey indicates bins falling within the 95\% percentiles.
These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
% \tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to alleviate confusion. I'm picturing a set of tables numbered with case counts in a top row and then a bottom row with sequence counts. This could also work better as a separate figure referenced from the Methods.}
	}
	\label{ABC}
\end{figure}

@@

Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (Figure \ref{recombination_tree}).
Neither method can identify where in the tree recombination occurred, but each full asterisk in Figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
Although the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur\lmc{stupid biostatistician here: is it?}.

@@

\subsection*{Role of population structure in phylodynamics}
\lmc{Detecting?} Population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.

@@

\section*{Methods}
\subsection*{Sequence data}
All MERS-CoV sequences were downloaded from GenBank.\lmc{accession numbers are given in Table X}.
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
Protein coding sequences were extracted and concatenated.
Sequences were annotated with available collection dates and hosts, designated as camel or human.
The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections.

>>>>

\usepackage{longtable}

@@

% symbols
% \newcommand{\chiSq}{\chi^{2}_{df}} %LM: ancient DNA? =P %GD: quite :)
% \newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
% \newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
% \newcommand{\dspr}{d_\mathrm{SPR}}

@@

% \date{\today}

@@

\section*{Introduction}
%% Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels. % AR - a virus isn't an infection and not all infections in the Arabian Peninsula
Middle East Respiratory Syndrome Coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans. % ? something like this ?
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
Whilst contact with camels is often reported, patients may not recall contact with %\lmc{other?}
any livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
Epidemiological modeling approaches have been used to look at traditional data sources of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches %\lmc{why, specifically?}
due to difficulties in linking cases into clusters in the absence of detailed information.
The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Often only sequence data can %\lmc{help}
help pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from %\lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}
outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
\citep{who_2017}.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.

%Unlike discrete trait analyses (DTA, or ``mugration'' models) %% AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
%that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) %[cite here, Philippe's bat rabies and Sam's avian flu?, others?]
\citep{faria_simultaneously_2013,global_consortium_for_h5n8_and_related_influenza_viruses_role_2016}, population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using Monte Carlo simulations we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.

@@

The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we %\lmc{estimate?}
estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels %\lmc{is ``allow'' the best word here? I understand you' rather not use ``biases'' or ``misleads'', though...}
the model cannot dismiss humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly %\lmc{maybe ``non-uniformly'' is redundant here} \gdc{disagree, the point I am trying to make is that the data are mostly human and that human data are further skewed in how they are sampled}
sampled human cases and are robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike the most simpler %\lmc{simpler?}
``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, structured population models possess the ability to estimate migration times.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{MERS-CoV seasonality.}
A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Grey area spanning October to February indicates the camel calving season.
B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (Figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
%Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
	}
	\label{seasonality}
\end{figure}

@@

Structured coalescent approaches clearly show humans to be a terminal host %\tbc{again, I don't know if this is the right term or not}
for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides %\lmc{insights?}
insight into epidemic behavior.
The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by a dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}).
We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
First, we simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from index cases will be sampled exceedingly rarely.
We record the $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses (left-hand panels in Figure \ref{ABC}).

We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%%\lmc{again?}
(nearly 10 000 out of 363 000 for both) of simulations fit for bias levels of 2.0 and 3.0, respectively.
This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate that a higher number of simulations generate MERS-CoV-like data at bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).%\lmc{I don't follow this argument...}
Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).

Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{ABC}).
The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
Our results suggest a large number of unobserved MERS index cases.%\lmc{does the definition of ``index case'' allow for ``multiple''?} \gdc{not in epi literature I don't think}
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In those sets of analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
	\caption{\textbf{Results of Monte Carlo simulations.}
Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced%\lmc{make sequences from large clusters more likely to be picked}
).
Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
Darker shade of grey indicates bins falling within the 95\% percentiles.
These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
% \tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to alleviate confusion. I'm picturing a set of tables numbered with case counts in a top row and then a bottom row with sequence counts. This could also work better as a separate figure referenced from the Methods.}
	}
	\label{ABC}
\end{figure}

@@

Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (Figure \ref{recombination_tree}).
Neither method can identify where in the tree recombination occurred, but each full asterisk in Figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
Although the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
This provides ideal conditions for co-infection with distinct genotypes, which is a pre-requisite for detectable RNA virus recombination to occur%\lmc{stupid biostatistician here: is it?} \gdc{fair comment actually, it's a pre-requisite for detectable recombination to occur usually}
.

@@

\subsection*{Role of population structure in phylodynamics}
%\lmc{Detecting?}
Understanding population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.

@@

\section*{Methods}
\subsection*{Sequence data}
All MERS-CoV sequences were downloaded from GenBank.%\lmc{accession numbers are given in Table X}.
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
Protein coding sequences were extracted and concatenated.
Sequences were annotated with available collection dates and hosts, designated as camel or human.
The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).

\begin{longtable}{ | l | l | l | l | l | }

  \caption{Strain names, accessions (where available), identified host and reported collection dates for MERS-CoV genomes used in this study.} \label{sequences} \\

  \hline & strain & accession & host & collection date \\ \hline
  \endfirsthead

  \multicolumn{5}{l}%
  {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\

  \hline & strain & accession & host & collection date \\ \hline
  \endhead

  \hline \multicolumn{5}{|r|}{{Continued on next page}} \\ \hline
  \endfoot

  \hline \hline
  \endlastfoot
